---
title: "Identifying DNA fragment that increases (Solutions)"
output: html_notebook
---
## Manipulating data tables

We will be using the tidyverse in this lab. So let's start by loading it.

```{r}
library(tidyverse)
```

Your lab conducted an experiment in which four different fragments of chromosome 21 were integrated into mice. The four parts are denoted with _141G6_, _152F7_, _230E8_ and _285E6_. The mice were bred resulting in dozens of  transgenic mice. The DNA fragment is not always inherted so some mice have the extra copy and others don't. We are interested in determining if any of these fragments result in an increase in weight, a characteristic associated with trisomic mice.

The data can be loaded like this.

```{r}
load("../rdas/mouse.rda")
```

Which loads the `dat` object into your environment:

```{r}
dat
```

The columns included in this table are the following:

  - _DNA_: Fragment of chromosome 21 integrated in parent mouse (1=141G6; 2=152F7; 3=230E8; 4=285E6).
  - _line_: Family line.
  - _tg_ - Whether the mouse contains the extra DNA (1) or not (0).
  - _sex_:  Sex of mouse (1=male; 0=female).
  - _age_: Age of mouse (in days) at time of weighing.
  - _weight_: Weight of mouse in grams, to the nearest tenth of a gram.
  - _bp_: Blood pressure of the mosue.
  - _cage_: Number of the cage in which the mouse lived

Let's start by comparing the weights of the no trisomic mice to the weights of mice with the other four fragmets. Determine which columns tells us the fragment of the mouse and determine the number of mouse in each group? Hint: use the _count_ function.

```{r}
dat %>% count(DNA)
```

Note that the names are 1, 2, 3, 4. Let's change these to the actual names
1=141G6; 2=152F7; 3=230E8; 4=285E6. Create a new column called `fragment` with the actual fragment names. Hint: Use the `recode` funtion. 

```{r}
dat <- mutate(dat, fragment = recode(DNA, 
                                "1"="141G6", 
                                "2"="152F7", 
                                "3"="230E8", 
                                "4"="285E6"))
```

Note that all the mice in our table have one of these names. However, we know that not all mouse have the fragments. Remember that not all inhereted the extra copy. Use `filter` and `count` to see how many mice in the `141G6` group have the extra copy.

```{r}
filter(dat, fragment == "141G6") %>% count(tg)
```

Now change the `fragment` column so that the mice that do not have the extra copy, have are called  `No trisomy`. Hint: use the `ifelse` function.

```{r}
dat <- dat %>% mutate(fragment = ifelse(tg == 0, "No trisomy", fragment)) 
```

Before we continue let's learn about the `n()` function. Note that we can perform the same as the `count()` function using `group_by()` and `n()`

```{r}
dat %>% group_by(DNA) %>% summarize(freq = n())
```

Now compute the average and standard error in each of the four groups and the control. Hint: Use `group_by` and `summarize`. 

```{r}
dat %>% group_by(fragment) %>% 
  summarize(average = mean(weight), se = sd(weight)/sqrt(n()))
```

Bonus: Is the above difference statistically signficant at the 0.05 level?

```{r}
lm(weight ~ tg, data = dat) %>% summary() %>% .$coef
```

This is what the dynamite plot would look like:

```{r, echo=FALSE}
dat %>% group_by(fragment) %>% 
  summarize(average = mean(weight), se=sd(weight)/sqrt(n())) %>%
  ggplot(aes(fragment, average)) + 
  geom_errorbar(aes(ymin = average - se, ymax = average+se), width = 0.25) +
  geom_bar(stat = "identity", width=0.5, fill=4, col = 1) +
  ylab("Outcome") +
  xlab("") 
```

Does this make sense? In the next section we demonstrate how data exploration allows us to detect a problem. We also learn how to make publication ready plots.

## Data visualization

### Data exploration to identify outliers

A problem with the summary statistics and the barplot above is that it only shows the average and we learn little about the distribution of the data. Use the _geom_boxplot_ function to show the five number summary. Do you see a problem? What is it?

```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot()
```

We know that a 1000 gram mice does not exist. In fact 100 grams is already a huge mouse. Use filter to show the data from the mice weighing more than 100 grams. 

```{r}
filter(dat, weight > 100) 
```

What are the weights?

```{r}
filter(dat, weight > 100) %>% pull(weight)
```

An unfortunate common practice is to use the number 999 to denote missing data. The recommended practice is to acutely type NA. Use the fitler function remove all the rows with these missing values, then remake the figure and recompute the averages and standard errors.

```{r}
dat <- filter(dat, weight != 999)

dat %>% ggplot(aes(fragment, weight)) + geom_boxplot()

dat %>% group_by(fragment) %>% 
  summarize(average = mean(weight), se = sd(weight)/sqrt(n()))
```

This makes much more sense.

## Data exploration to help answer scientific question

Let us start by creating a boxplot of weight vs fragments:

```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot()
```

We prefer showing the control data first. To achieve this we need to 
let R know that `No trisomy` is the reference. We can do this by converting the variable into a factor and using the `relevel` function. Like this:

```{r}
dat <- mutate(dat, fragment = factor(fragment)) %>% 
       mutate(fragment = relevel(fragment, ref = "No trisomy"))

dat %>% ggplot(aes(fragment, weight)) + geom_boxplot()
```

Notice that the boxplot do show evidence of asymetry. Use the `geom_point()` function to add the points to the boxplot.

```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot() + 
  geom_point()
```

The points are all cluster together and its hard to see them. We can use `geom_jitter` instead of `geom_point` to fix this:

```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot() + 
  geom_jitter()
```

What did `geom_jitter` do? Let's make the spread smaller and alpha-blend the data points

```{r}
dat %>% ggplot(aes(fragment, weight)) +
  geom_boxplot() + 
  geom_jitter(width = 0.1, alpha = 0.5)
```

Do you see anything interesting? Note that an unexpected result is revealed, for some groups it seems like we have a bimodal distribution. We can use histograms to assess this. For example, here is the histogram for the control group:

```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_histogram()
```

To avoid the warning we can add the argument  `bindwith = 1` and to make it nicer looking we can add `color = "black"`.

```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_histogram(binwidth = 1, color = "black")
```

There appears to be evidence for two modes. We can also see this using the smooth density estimator using `geom_density`:

```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_density()
```

Note that this is similar  to the histogram above. Now let's make it nice:

```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_density(fill = 1, alpha = 0.5) + 
  xlim(c(19,45))
```

We only looked at the distribution of `weight` in the control group. We can use ridge plots to compare the distribution across all groups. To use ridge plots we need to install and load the `ggridges` functions.

```{r}
library(ggridges)
dat %>% 
  ggplot(aes(weight, fragment)) +
  geom_density_ridges() 
```

To avoid the overlap you can use:

```{r}
library(ggridges)
dat %>% 
  ggplot(aes(weight, fragment)) +
  geom_density_ridges(scale = 0.9) 
```

We see that for most groups we see two modes. What could explain this?

One variable to consider is sex. We can remake the boxplot with points above but this time use color to distinguish the data from males and females.

```{r}
dat %>% 
  ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5)
```

A problem here is that sex is a number! It should be a character or a factor. Change it to a character vector using `mutate` then remake the plot

```{r}
dat <- mutate(dat, sex = ifelse(sex == 1, "Male", "Female"))

dat %>% ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5)
```

Now that we see that males and females have different distributions, the question arises if the fragment effects are different for males and females. We can explore this by remaking the plot for each sex separately. This can be achieved using the the faceting.

```{r}
dat %>% ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  facet_grid(.~sex)
```

A quick pause to describe themes. `ggplot` provides the options of changing the general look of plots. There is an entire package, __ggthemes__, dedicated to providing different themes. Here we use the black and white theme:

```{r}
dat %>% ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  facet_grid(.~sex) + 
  theme_bw()
```

This plot is quite informative. First it shows strong evidence that the 152F7 has an effect on weight, especially on females.

### Cage effect?

Are males and females caged together?

```{r}
with(dat, table(cage, sex))
```

Is cage counfounded with fragment? Make boxplots for the females.

```{r}
dat %>% 
  filter(sex == "Female") %>%
  mutate(cage = factor(cage)) %>%
  ggplot(aes(cage, weight, color = fragment)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  theme_bw()
```

When order by meaningful values.

```{r}
dat %>% 
  filter(sex == "Female") %>%
  mutate(cage = factor(cage)) %>%
  mutate(cage = reorder(cage, weight, median)) %>%
  ggplot(aes(cage, weight, color = fragment)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  theme_bw()
```

Take a closer look. Compare just controls and 152F7. Are trisomic mice higher within each cage? Is what we are seeing really a cage effect?

```{r}
dat %>% 
  filter(sex == "Female" & fragment %in% c("No trisomy", "152F7")) %>%
  mutate(cage = factor(cage)) %>%
  mutate(cage = reorder(cage, weight, median)) %>%
  ggplot(aes(cage, weight, color = fragment)) +
  geom_boxplot(aes(color = fragment), width = 0.5) + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  theme_bw()
```

### Confounding

For the female mice compute the correlation between blood pressure and weight.

```{r}
dat %>% filter(sex == "Female") %>%
  summarize(cor(weight, bp))
```

The correlation is negative. Does this make sense? Confirm with a plot that there are not outliers driving this result.

```{r}
dat %>% filter(sex == "Female") %>%
  ggplot(aes(weight, bp)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method="lm", se=FALSE)
```


The plot does confirm that higher weight mice have, on average slightly lower blood pressure. But we do see clusters. What could these be? Use color to decipher what cuases the clustering.

```{r}
dat %>% filter(sex == "Female") %>%
  ggplot(aes(weight, bp, color = fragment)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method="lm", se=FALSE)
```

Now use faceting to plot the scatter plot for each fragment separately.

```{r}
dat %>% filter(sex == "Female") %>%
  ggplot(aes(weight, bp)) + 
  geom_point(alpha = 0.5) + 
  facet_wrap(fragment ~ .)
```

We note that the correlation appears positive in each group. Use `group_by` and `summarize` to corroborate this.

```{r}
dat %>% filter(sex == "Female") %>%
  group_by(fragment) %>%
  summarize(cor = cor(weight, bp))
```

Note: Cases in which the overall correlation is the opposite sign as when looking strata is referred to as Simpsons Paradox.


