---
title: "Data Visualization in Practice (Workshop)"
output: github_document
---

Today's workshop is an adaptation of [Chapter 9 of Introduction to Data Science: Data Analysis and Prediction Algorithms with R](https://rafalab.github.io/dsbook/gapminder.html). We will demonstrate the usefulness of __ggplot2__ with a case study on insights on poverty popularized by [Hans Rosling](https://en.wikipedia.org/wiki/Hans_Rosling), the co-founder of the [Gapmider Foundation](http://www.gapminder.org/). Specifically, today we will use data to attempt to answer the following two questions:

1. Is it a fair characterization of today's world to say it is divided into western rich nations and the developing world in Africa, Asia, and Latin America? 
2. Has income inequality across countries worsened during the last 40 years? 

To answer these questions, we will be using the `gapminder` dataset provided in __dslabs__. This dataset was created using a number of spreadsheets available from the Gapminder Foundation. You can access the table like this:

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(dslabs)
data(gapminder)
ds_theme_set()
gapminder %>% as_tibble()
```

### Hans Rosling's quiz 

Let's start by testing our knowledge regarding differences in child mortality across different countries. For each of the pairs below, which country do you think had the highest child mortality rates in 2015? Which pairs do you think are most similar?

1. Sri Lanka or Turkey
2. Poland or South Korea
3. Malaysia or Russia
4. Pakistan or Vietnam
5. Thailand or South Africa

When answering these questions without data, the non-European countries are typically picked as having higher child mortality rates: Sri Lanka over Turkey, South Korea over Poland, and Malaysia over Russia. It is also common to assume that countries considered to be part of the developing world: Pakistan, Vietnam, Thailand, and South Africa, have similarly high mortality rates. 

To answer these questions __with data__, we can use __dplyr__. Use the `gapminder` dataset to answer the questions above:

```{r, message=FALSE}
pairs <- list(c("Sri Lanka","Turkey"),
     c("Poland","South Korea"),
     c("Malaysia","Russia"),
     c("Pakistan","Vietnam"),
     c("Thailand","South Africa"))

for(pair in pairs)
{
  p = gapminder %>% 
  filter(year == 2015 & country %in% pair) %>% 
  select(country, infant_mortality)
  
  print(p)
}
```

It turns out that when Hans Rosling gave this quiz to educated groups of people, the average score was less than 2.5 out of 5, worse than what they would have obtained had they guessed randomly. This implies that more than ignorant, we are misinformed. Let's see how data visualization helps to inform us.

## Scatterplots

The reason for this stems from the preconceived notion that the world is divided into two groups: the western world (Western Europe and North America), characterized by long life spans and small families, versus the developing world (Africa, Asia, and Latin America) characterized by short life spans and large families. But do the data support this dichotomous view?

The necessary data to answer this question is also available in our `gapminder` table. Using our newly learned data visualization skills, we will be able to tackle this challenge.

Start by creating a scatterplot of life expectancy versus fertility rates (average number of children per woman) on 1952

```{r, warning=FALSE}
filter(gapminder, year == 1962) %>%
  ggplot(aes(fertility, life_expectancy)) +
  geom_point()
```

Most points fall into two distinct categories: 

1. Life expectancy around 70 years and 3 or fewer children per family.
2. Life expectancy lower than 65 years and more than 5 children per family. 

As we learned today, let's use color to codify continent:

```{r}
filter(gapminder, year == 1962) %>%
  ggplot( aes(fertility, life_expectancy, color = continent)) +
  geom_point() 
```

There is some truth on "the West versus developing world" view back in 1962. Is this still true?

## Faceting
Let's plot the 2012 data in the same way we did for 1962. Recall one of the principles discussed today: __use common axes__. In __ggplot2__, we can achieve this by __faceting__ variables with `facet_grid`. Here is an example:

```{r, warning=FALSE, out.width="100%"}
filter(gapminder, year%in%c(1962, 2012)) %>%
  ggplot(aes(fertility, life_expectancy, col = continent)) +
  geom_point() +
  facet_grid(continent~year)
```

This is just an example and more than what we want, which is simply to compare 1962 and 2012. In this case, we only want a window for 1962 and another for 2012. Fix the code from above to achieve this:

```{r, warning=FALSE, out.width="100%", fig.height=3}
filter(gapminder, year%in%c(1962, 2012)) %>%
  ggplot(aes(fertility, life_expectancy, col = continent)) +
  geom_point() +
  facet_grid(. ~ year)
```

This plot clearly shows that the majority of countries have moved from the _developing world_ cluster to the _western world_ one. In 2012, the western versus developing world view no longer makes sense. This is particularly clear when comparing Europe to Asia, the latter of which includes several countries that have made great improvements. 

### `facet_wrap`

To explore how this transformation happened through the years, we can make the plot for several years. For example, we can add 1970, 1980, 1990, and 2000. Use `facet_wrap` to achieve this:

```{r, out.width="100%"}
years <- c(1962, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
gapminder %>% 
  filter(year %in% years & continent %in% continents) %>%
  ggplot( aes(fertility, life_expectancy, col = continent)) +
  geom_point() +
  facet_wrap(~year) 
```

This plot clearly shows how most Asian countries have improved at a much faster rate than European ones.

## Time series plots

The visualizations above illustrate that the data no longer supports the western versus developing world view. Now a few new questions emerge:

- Which countries are improving more and which ones less?
- Was the improvement constant during the last 50 years or was it more accelerated during certain periods?

For a closer look that may help answer these questions, we introduce _time series plots_. Time series plots have time in the x-axis and an outcome or measurement of interest on the y-axis. For example, here is a trend plot of United States fertility rates:

```{r, warning=FALSE}
gapminder %>%
  filter(country == "United States") %>%
  ggplot(aes(year, fertility)) +
  geom_point()
```

Note that the trend is not linear. We see a sharp decrease during the 1960s and 1970s to below 2 and then a slow increase and subsequent  plateau around 2.

When the points are regularly and densely spaced, as they are here, we create curves by joining the points with lines, to convey that these data are from a single series, here a country. To do this, we use the `geom_line` function instead of `geom_point`. Try it for yourself:

```{r, warning=FALSE}
gapminder %>%
  filter(country == "United States") %>%
  ggplot(aes(year, fertility)) +
  geom_line()
```

This is particularly helpful when we look at two countries. If we subset the data to include two countries, one from Europe and one from Asia, then adapt the code above:

```{r, warning=FALSE, message=FALSE}
countries <- c("South Korea","Germany")

gapminder %>% filter(country %in% countries) %>%
  ggplot(aes(year,fertility)) +
  geom_line()
```

Any comments? Is this the plot we want? How can we fix it? Try it for yourself:

```{r, warning=FALSE, message=FALSE}
countries <- c("South Korea","Germany")

gapminder %>% 
  filter(country %in% countries) %>%
  ggplot(aes(year, fertility, color=country)) +
  geom_line()
```

This figure clearly shows how South Korea's fertility rate dropped drasticallly during the 1960 and 70s, and by 2000 had a similar rate to that of Germany.

### Labels instead of legends

For trend plots we recommend labeling the lines rather than using legends since the viewer can quickly see which line is which country. This suggestion actually applies to most plots: labeling is usually preferred over legends.

Use the code from above to create a time-series plot of life expectancy for the same countries:

```{r, warning=FALSE, message=FALSE}
countries <- c("South Korea","Germany")

gapminder %>% 
  filter(country %in% countries) %>%
  ggplot(aes(year, life_expectancy, color=country)) +
  geom_line()
```

We demonstrate how we can do this using the life expectancy data. We define a data table with the label locations and then use a second mapping just for these labels:

```{r}
labels <- data.frame(country = countries, 
                     x = c(1975,1965), 
                     y = c(60,72))

gapminder %>%
  filter(country %in% countries) %>%
  ggplot(aes(year, life_expectancy, col = country)) +
  geom_line() +
  geom_text(data = labels, aes(x, y, label = country), size = 5) +
  theme(legend.position = "none")
```

The plot clearly shows how an improvement in life expectancy followed the drops in fertility rates. In 1960, Germans lived 15 years longer than South Koreans, although by 2010 the gap is completely closed. It exemplifies the improvement that many non-western countries have achieved in the last 40 years.

## Data transformations

Let us now address the second question: Has income inequality across countries worsened during the last 40 years? Would you agree with the following statement: __Poor countries getting poorer and richer countries are getting richer__. The `gapminder` data table includes a column with the countries' gross domestic product (GDP). GDP measures the market value of goods and services produced by a country in a year. The GDP per person is often used as a rough summary of a country's wealth. Here we divide this quantity by 365 to obtain the more interpretable measure _dollars per day_.  Using current US dollars as a unit, a person surviving on an income of less than $2 a day is defined to be living in _absolute poverty_. We add this variable to the data table:

```{r}
gapminder <- gapminder %>%  mutate(dollars_per_day = gdp/population/365)
```

The GDP values are adjusted for inflation and represent current US dollars, so these values are meant to be comparable across the years. Of course, these are country averages and within each country there is much variability. All the graphs and insights described below relate to country averages and not to individuals.

### Log transformation

Here is a histogram of per day incomes from 1970:

```{r}
past_year <- 1970
gapminder %>%
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day)) +
  geom_histogram(binwidth = 1, color = "black")
```

From this plot we see that most countries have an average daily income below \$10 a day. Therefore, it might be more informative to quickly be able to see how many countries have average daily incomes of about $1 (extremely poor), \$2 (very poor), \$4 (poor), \$8 (middle), \$16 (well off), \$32 (rich), \$64 (very rich) per day. These changes are multiplicative and log transformations convert multiplicative changes into additive ones: when using base 2, a doubling of a value turns into an increase by 1. Use the code from above to create a new figure with a log2 transformation:

```{r}
##
gapminder %>%
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(log2(dollars_per_day))) +
  geom_histogram(binwidth = 1, color = "black")

##
gapminder %>%
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2")
```

## Visualizing multimodal distributions

The histogram above suggests that the 1970 country income distribution has two modes: one at about 2 dollars per day (1 in the log 2 scale) and another at about 32 dollars per day (5 in the log 2 scale). This _bimodality_ is consistent with a dichotomous world made up of countries with average incomes less than $8 (3 in the log 2 scale) a day and countries above that.

## Comparing multiple distributions with boxplots and ridge plots

The histogram above provides evidence of _bimodality_. However, the histogram does not show us if the two groups of countries are _west_ versus the _developing_ world. Use `reorder` to create a scatterplot of `dollars_per_day` versus `region`:

```{r}
gapminder %>%
  filter(year == past_year & !is.na(gdp)) %>%
  mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%
  ggplot(aes(dollars_per_day, region)) +
  geom_point() +
  scale_x_continuous(trans = "log2")
```

We can already see that there is indeed a "west versus the rest" dichotomy: we see two clear groups, with the rich group composed of North America, Northern and Western Europe, New Zealand and Australia. We define groups based on this observation:

```{r}
gapminder <- gapminder %>%
  mutate(group = case_when(
    region %in% c("Western Europe", "Northern Europe","Southern Europe",
                    "Northern America",
                  "Australia and New Zealand") ~ "West",
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region %in% c("Caribbean", "Central America",
                  "South America") ~ "Latin America",
    continent == "Africa" &
      region != "Northern Africa" ~ "Sub-Saharan",
    TRUE ~ "Others"))
```

We turn this `group` variable into a factor to control the order of the levels:

```{r}
gapminder <- gapminder %>%
  mutate(group = factor(group, levels = c("Others", "Latin America",
                                          "East Asia", "Sub-Saharan",
                                          "West")))
```

In the next section we demonstrate how to visualize and compare distributions across groups.

### Boxplots

Using the `group` variable that we defined above, create a boxplot of `dollars_per_day` versus `group`:

```{r}
p <- gapminder %>%
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(group, dollars_per_day)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```

Recall our first principle, show the data!

```{r}
p + geom_jitter(alpha = 0.5, width = 0.10)
```

### Ridge plots

Showing each individual point does not always reveal important characteristics of the distribution. Although not the case here, when the number of data points is so large that there is over-plotting, showing the data can be counterproductive. Boxplots help with this by providing a five-number summary, but this has limitations too. For example, boxplots will not permit us to discover bimodal distributions. To see this, note that the two plots below are summarizing the same dataset:

```{r, warning=FALSE, out.width="100%", message = FALSE, echo=FALSE, fig.height=3}
set.seed(1987)
z <- sample(c(0,1), 1000, replace = TRUE, prob = c(0.25, 0.75))
x <- rnorm(100)*z + rnorm(100, 5)*(1 - z)
p1 <- qplot(x, geom = "density", fill = 1, show.legend=FALSE, alpha = 0.2) +
  scale_x_continuous(limits=c(-4,8.5))
p2 <- qplot("", x, geom="boxplot")
gridExtra::grid.arrange(p1, p2, nrow = 1)
```

In cases in which we are concerned that the boxplot summary is too simplistic, we can show stacked smooth densities or histograms. We refer to these as _ridge plots_. The package __ggridges__ provides a convenient function for doing this. Here is the income data shown above with boxplots but with a _ridge plot_:

```{r, message=FALSE}
library(ggridges)
p <- gapminder %>%
  filter(year == past_year & !is.na(dollars_per_day)) %>%
  ggplot(aes(dollars_per_day, group)) +
  scale_x_continuous(trans = "log2")
p  + geom_density_ridges()
```

Note that we have to invert the `x` and `y` used for the boxplot. A useful `geom_density_ridges` parameter is `scale`, which lets you determine the amount of overlap, with `scale = 1` meaning no overlap and larger values resulting in more overlap. Here is an example:

```{r message=FALSE, warning=FALSE}
p + geom_density_ridges(scale = 2)
```

If the number of data points is small enough, we can add them to the ridge plot using the following code:

```{r, message=FALSE}
p + geom_density_ridges(jittered_points = TRUE)
```

By default, the height of the points is jittered and should not be interpreted in any way. To show data points, but without using jitter we can use the following code to add what is referred to as a _rug representation_ of the data.

```{r, message=FALSE}
p + geom_density_ridges(jittered_points = TRUE,
                        position = position_points_jitter(height = 0),
                        point_shape = '|', point_size = 3,
                        point_alpha = 1, alpha = 0.7, scale = 1)
```

### Example: 1970 versus 2010 income distributions

Data exploration clearly shows that in 1970 there was a "west versus the rest" dichotomy. But does this dichotomy persist? Let's use `facet_grid` see how the distributions have changed. To start, we will focus on two groups: the west and the rest. We make four histograms.

```{r}
past_year <- 1970
present_year <- 2010
years <- c(past_year, present_year)
gapminder %>%
  filter(year %in% years & !is.na(gdp)) %>%
  mutate(west = ifelse(group == "West", "West", "Developing")) %>%
  ggplot(aes(dollars_per_day)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2") +
  facet_grid(year ~ west)
```















